dispRity manual

### 2016-12-29 doc/dispRity-manual.Rmd
Partially updated manual

`dispRity` is a package for calculating disparity in `R`. It allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.

@@

Because this package is aimed to be multidisciplinary, many names or terms used in this tutorial might be non-familiar to certain fields.
Here is a list of what are the exact meaning of these term:

-   **Ordinated space**: it designates the mathematical multidimensional object studied here.
    In morphometrics, this one is often referred as being the morphospace.
    However it can also be the the cladisto-space for cladistic data or the eco-space in ecology, etc.
    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the ordinated space (often – but not necessarily - > 3!) and the rows represent the elements within this space.

-   **Elements**: it designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...

-   **Dimensions**: it designates the columns of the ordinated space. The dimensions can also be referred to as axis or principal components.

-   **Series**: it designates sub-samples of the ordinated space.
    Basically a series contain the same number of dimensions as the morphospace but may contain a smaller number of elements.
    For example, if our ordinated space is composed of birds and mammals (i.e. the elements) and 50 dimensions, we can create two series of just mammals or birds as elements (but with the same 50 dimensions) to look at the difference in disparity between the two clades.

@@

Note that we use the `release` branch here which is version `0.2`.
For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
This package depends mainly on the `ape` package and uses functions of several other (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree` and `snow`).

@@

In this tutorial we are going to use a subset of the ordinated cladistic data from @beckancient2014 that contains 50 taxa (elements) ordinated using their cladistic distance (i.e. the distance between their discrete morphological characters).
Note that this data is more oriented towards palaebiology analysis but that it can apply to other disciplines.
Please refer to the [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes covering some specific example.

@@

## Loading the phylogeny
data(BeckLee_tree)
plot(BeckLee_tree, cex = 0.8) ; nodelabels(cex = 0.7) ; axisPhylo(root = 140)
```

@@

```{r, fig.width=6, fig.height=6}
## Splitting the data
sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
    model = "acctran", time = rev(seq(from=0, to=130, by=15)),
    FADLAD = BeckLee_ages)

## Bootstrapping the data
bootstrapped_data <- boot.matrix(sliced_data, 100)

## Calculating disparity
median_centroids <- dispRity(bootstrapped_data, metric = c(median, centroids))

## Summarising the results
summary(median_centroids)
plot(median_centroids)

## Testing the effect of time on disparity
summary(test.dispRity(median_centroids, test = aov, comparisons = "all"))
```

@@

```{r, eval = FALSE}
str(median_centroids)
## That's a more than 5000 lines of output!
```

@@

@@

## Summarising it using the S3 method print.dispRity:
median_centroids
## This displays some basic information about the object content
```

@@

```{r, eval=FALSE}
## Displaying the full object
print(median_centroids, all = TRUE)
## But that's more than 155000 lines!
```

Some utility functions such as `sort.dispRity`, `get.dispRity` or `extract.dispRity` allows to access or reorder to some specific content of the object:

```{r}
## Reordering the series (from the present to the past)
sort.dispRity(median_centroids, decreasing = TRUE)

## Extracting some specific series from the disparity object
get.dispRity(median_centroids, what = c(1,4))

## Extracting the observed disparity
extract.dispRity(median_centroids, observed = TRUE)

## The distribution of the bootstrapped disparity scores for the first series
str(extract.dispRity(get.dispRity(median_centroids, what = 1),
    observed = FALSE))
```

@@

This package aims to be a *modular package*, that is, it is designed to be fairly easily modified by users.
The modular aspect allows the package users to use several implemented tools to help them creating personalised functions.
For example, the `dispRity` function intake a `metric` argument designating how disparity should be calculated.
This argument can take some already implemented functions such as `mean` but it can also intake some completely new and revolutionary ones such as the sum divided by length:

@@

## A function for measuring the sum divided by the length (the mean!)
sum.divide.by.length <- function(X) sum(X)/length(X)

## Disparity measured as the mean divided by the length
summary(dispRity(sliced_data, metric = sum.divide.by.length))
## That is a bit like reinventing the wheel...
```

@@

In the following section, different functionalities are detailed.
The functions are presented in a "classic" pipelined study fashion for convenience (i.e. splitting the data, bootstrapping it, calculating disparity, testing differences).
Note however that most of these functions can be used in any order and that some steps can of course be skipped and that the package contains many more additional functions!

@@

The `cust.series` function allows to create these series according to factors determined by the user.
In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (i.e. all the living mammals and their direct ancestor) and the stem mammals (i.e. all the mammals that have no related living descendants).

```{r}
## We want to separate the species around the node 71 (see phylogeny above).
## All the descendant of these node are crown and all the ancestors are stem.
crown <- extract.clade(BeckLee_tree, node = 71)$tip.label
stem <- drop.tip(BeckLee_tree, tip = crown)$tip.label
## We then have to feed this information in a data frame with one column
factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem), "Group")))

## We then can use the customised series function to create the two series
## i.e. the two regions of the ordinated space
crown_stem <- cust.series(BeckLee_mat50, factors)
```

The results of this function (like many others presented below) output a `dispRity` object:

```{r}
crown_stem

## This object contains three elements
names(crown_stem)
## With "data" being the list sub-matrices
str(crown_stem$data)
## "elements" being the list of taxa in the original ordinated matrix
str(crown_stem$elements)
## and finally, "series" containing information on the series type (custom)
## and names (crown and stem)
crown_stem$series
```

@@

```{r}
## Adding all mammals to the a new series (along with crown and stem as before)
factors2 <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem)),
    rep("Mammalia", Ntip(BeckLee_tree))), ncol = 2,
    dimnames = list(c(crown, stem), c("Group", "Class")), byrow = FALSE))

## Creating the three series: Group.stem/crown and Class.Mammalia
cust.series(BeckLee_mat50, factors2)
```

@@

Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
The `time.series` function allows to create series that contain all the elements present at specific points or during specific periods in time.
This functions needs as input an ordinated space and a matching phylogenetic tree.
Two types of time series can be performed by using the `method` option:

@@

For the time-slicing method details see @GuillermeSTD.
For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:

@@

```{r}
## Generating three time bins containing the data present every 40 Ma
time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = c(120, 80, 40, 0))

In this example, the taxa where split inside each time-bin according to their age.
Note however that some taxa might span between some time bins and should be included in more than one.
This is possible by providing a table containing the first and last occurrence data:

```{r}
## Generating three time bins containing the data present every 40 Ma, including
## taxa that might span between time bins.
(time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))

This generated indeed a list of 3 sub matrices.
Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:

@@

For the time-slicing method (`method = continuous`), the idea is really similar.
This option intakes a matrix that contains the ordinated distance between taxa *and* nodes and an assumed evolutionary model via the `model` argument:

@@

```{r}
## Generating four time slices every 40 million years assuming a gradual
## evolutionary model
(time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages))
## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
    model = "gradual", time = 4)
```

@@

Once we obtain our different series, it is possible to bootstrap and rarefy it (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare slices with the same number of elements to get rid of eventual sampling problems.
The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
The default options will bootstrap the matrix 1000 times without rarefaction.
The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option (see below).

```{r}
## Bootstrapping a matrix
boot.matrix(data = BeckLee_mat50)
```

This function allows to control the bootstrap algorithm through the `boot.type` argument.
Currently two algorithms are implemented:

@@

This function also allows to rarefy the data using the `rarefaction` argument.
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in the dataset to a minimum of 3 elements).
It can also be set to a fix `numeric` value (or a set of `numeric` values).
Finally, one last argument, `rm.last.axis` allows to remove a certain amount of dimensions (or axis) for the ordinated space.
This can be logical argument where `FALSE` (default) will not remove any dimension and `TRUE` will remove the last dimensions that bear up to 5% of the total ordinated space’s variance.
Note that it can also intake a `numerical` value designating the proportion of axis to keep.

@@

## And removing the last axis (default)
boot.matrix(BeckLee_mat50, rm.last.axis = TRUE)

## Or with a fix value (50%)
boot.matrix(BeckLee_mat50, rm.last.axis = 0.5)
```

Of course, one could be interested in directly supplying the sub-sampled matrices generated above directly to this function.
In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `cust.series` or `time.series` functions.

```{r}
## Bootstrap and full rarefaction on the crown/stem series
(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))

## Bootstrap on the time binning/slicing series
(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
```

@@

This function is a modular function that allows to simply (and quickly!) calculate disparity from a matrix.

Because disparity can be measured in many ways, this function is a tool to measure disparity *as defined by the user*.
In fact, the `dispRity` function intakes two main arguments: the data and the disparity metric.
The disparity metric is a function or a set of functions that summarises the ordinated matrix to a single value that represents, in our example, the diversity of morphologies.

@@

```{r}
dispRity(BeckLee_mat50, metric = mean)
## This defines disparity as the mean value of the morphospace
## (using base::mean).

## It is also possible to combine multiple functions:
dispRity(BeckLee_mat50, metric = c(sum, variances))
## Defining disparity as the sum (base::sum) of the variances
## (dispRity::variances) of each dimension of the morphospace.

dispRity(BeckLee_mat50, metric = c(prod, centroids))
## For the product (base::prod) of the distances between each elements and the
## centroid of the morphospace (dispRIty::centroids).

## Or user defined ones:
total.range <- function(X) abs(range(X)[1]-range(X)[2])
dispRity(BeckLee_mat50, metric = c(total.range, centroids))
## For the range (total.range, user defined) of the distances between each
## elements and the centroid of the morphospace (dispRity::centroids).

## Or more complex ones:
vars1 <- dispRity(BeckLee_mat50, metric = c(sd, variances, var))
## For the standard deviation (stats::sd) of the variance (dispRity::variances)
## of each column of the variance/covariance matrix (stats::var).

## Note that the order of the function is not important since the levels of
## each function are automatically detected by the dispRity algorithm.
vars2 <- dispRity(BeckLee_mat50, metric = c(variances, var, sd))
all(summary(vars1) == summary(vars2))
```

@@

In these examples we used only simple ordinated matrix but of course, it might be more interesting to directly use the `dispRity` objects we generated in the steps above.
Thus we can calculate the bootstrapped and rarefied disparity in each sub-sample of the morphospace.
Here we are going to define disparity as being the median distance between each element and each dimensions centroids.

@@

## Disparity through time:
disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))

Note that the computational time was longer for the rarefied data (`crown_stemBS`) since it had to calculate the disparity on all the rarefied matrix, each bootstrapped 100 times.
For both the `boot.matrix` and the `dispRity` functions, it is possible to increase computational speed in some cases using the `parallel` option (see each function’s manuals).

@@

This function is a S3 function (`summary.dispRity`) allowing to summarize the content of `dispRity` objects that contain disparity calculations.
This function intakes a `dispRity` object plus various options namely the `quantile` values for the confidence intervals levels; the `cent.tend` for the central tendency to use for summarising the results and two *visual* options which are whether to recall the `dispRity` options and how much digits are wanted in the results.

```{r}
summary(disp_time_binsBS)
## By default, the 50 and 95 quantiles and the mean are calculated.
## Note that the function also displays the observed disparity
## (non-bootstrapped).

## These arguments can be changed easily as follow
summary(disp_time_binsBS, quantile = 88, cent.tend = sd, rounding = 1)

@@

### `tree.age`

@@

Developments
============

As stated at the start of the demo, this version 0.2 is still in development and many parts are missing.
Here are the new functionalities that will be implemented in further versions.

More user defined functions
---------------------------

I intend also develop functions to help users to develop their own algorithms for the bootstrap method (via `make.boot`) or the evolutionary models (via `make.model`).
Both functions will provide similar testing as the `make.metric` function.

Suggestions?
------------

If you have any extra suggestions or comments on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).

>>>>

`dispRity` is a package for calculating disparity in `R`. It allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
<!-- Add version and table of content here! -->

@@

Because this package is aimed for multidisciplinary purpose, some names or terms used in this tutorial might be non-familiar to certain fields.
Here is a list of what are the exact meaning of these term:

-   **Ordinated space**: designates the mathematical multidimensional object that will be analysed with this package.
    In morphometrics, this one is often referred as being the morphospace.
    However it can also be the the cladisto-space for cladistic data or the eco-space in ecology, etc.
    In practice, this term designates an ordinated matrix where the columns represent the dimensions of the space (often – but not necessarily - > 3!) and the rows represent the elements within this space.

-   **Elements**: designates the rows of the ordinated space, elements can be either taxa, field sites, countries, etc...

-   **Dimensions**: designates the columns of the ordinated space. The dimensions can also be referred to as axis or principal components in the case of the ordinated space obtained from a PCA.

-   **Series**: it designates sub-samples of the ordinated space.
    Basically a series contain the same number of dimensions as the ordinated space but may contain a smaller subset of elements.
    For example, if our ordinated space is composed of birds and mammals (i.e. the elements) and 50 dimensions, we can create two series of just mammals or birds as elements (but with the same 50 dimensions) to look at the difference in disparity between the two clades.

@@

Note that we use the `release` branch here which is version `0.2`. <!-- add version !-->
For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
This package depends mainly on the `ape` package and uses functions of several other (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, and `RCurl`).

@@

In this tutorial we are going to use a subset of the ordinated cladistic data from @beckancient2014 <!-- fix citation -->
that contains 50 taxa (elements) ordinated using their cladistic distance (i.e. the distance between their discrete morphological characters).
Note that this data is more oriented towards palaebiology analysis but that it can apply to other disciplines.
Please refer to the [GitHub page](https://github.com/TGuillerme/dispRity) for other vignettes covering some specific example.

@@

## Loading the phylogeny
data(BeckLee_tree)
plot(BeckLee_tree, cex = 0.8) ; axisPhylo(root = 140) ; nodelabels(cex = 0.5)
```
<a name="mammal_tree"></a>

@@

```{r, fig.width=6, fig.height=6}
## Splitting the data into time series
sliced_data <- time.series(BeckLee_mat99, BeckLee_tree, method = "continuous",
    model = "acctran", time = rev(seq(from=0, to=130, by=15)),
    FADLAD = BeckLee_ages)

## Bootstrapping the data to measure its variance
bootstrapped_data <- boot.matrix(sliced_data, 100)

## Calculating disparity as the median distance from each element to the groups centroids
median_centroids <- dispRity(bootstrapped_data, metric = c(median, centroids))

## Summarising the disparity results
summary(median_centroids)
plot(median_centroids)

## Testing the effect of time on disparity (note that using an ANOVA violates some assumptions!)
summary(test.dispRity(median_centroids, test = aov, comparisons = "all"))
```

@@

```{r, eval = FALSE}
str(median_centroids)
## That's a more than 72 lines of lists of lists of matrices!
```

@@

<!-- TODO: change to html link -->

@@

## Summarising it using the S3 method print.dispRity
median_centroids
```

@@

```{r, eval=FALSE}
## Displaying the full object
print(median_centroids, all = TRUE)
## But that's more than ~ 2500 lines on a 13 inch laptop screen!
```

The package provides some utilities functions directly associated to the `dispRity` object manipulation such as `sort.dispRity`, `extract.dispRity` or `sort.dispRity` (see the full list in the [`dispRity` utilities](#dispRity.utilities) section).
These functions come in handy for modifying the `dispRity` object without having to go into it's structure!

@@

This package aims to be a *modular package*.
Its architecture is designed to be fairly easily modified by users.
The modular aspect allows the package users to use several implemented tools to help them creating personalised functions.
For example, the `dispRity` function intake a `metric` argument that is directly the function for calculating disparity should.
This argument can take some already implemented functions such as `mean` but it can also intake some completely new ones such as the sum divided by length (even though this example sounds like reinventing the wheel):

@@

## A function for measuring the sum divided by the length
sum.divided.by.length <- function(X) sum(X)/length(X)

## Disparity measured as the mean divided by the length
summary(dispRity(sliced_data, metric = sum.divided.by.length))
```

@@

In the following section, different functionalities of the package are detailed.
The functions are presented in a "classic" pipelined study fashion for convenience (i.e. splitting the data, bootstrapping it, calculating disparity, testing differences).
Note however that most of these functions can be used in any order and that some steps can of course be skipped and that the package contains many more additional functions!

@@

The `cust.series` function allows to create these series according to one or more factors determined by the user.
In this example, we can split the matrix based on the phylogeny and separate to different groups of taxa: the crown mammals (all the living mammals and their direct ancestor) and the stem mammals (all the mammals that have no related living descendants).
In order to do so, we need to separate the tree into two subtrees at the node 71 (see [phylogeny above](#mammal_tree)).

```{r}
## Selecting the crown and stem mammals
crown <- extract.clade(BeckLee_tree, node = 71)$tip.label
stem <- drop.tip(BeckLee_tree, tip = crown)$tip.label
```

We can then feed these lists of species to the `cust.series` function after transforming it into a factorial table:

```{r}
## Creating the factor table
series_factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem), "Group")))
head(series_factors)

## Creating the two series
crown_stem <- cust.series(BeckLee_mat50, series_factors)
```

The results of this function (like many others presented below) outputs a `dispRity` object:

```{r}
crown_stem
```

The two series represent two groups occupying specific regions of the mammalian morphospace.
The `dispRity` object itself always contains the original morphospace (`$matrix`) as well as information on which taxa compose the series (`$series`):

```{r}
## This object contains three elements
names(crown_stem)
## With "matrix" being the original matrix (the ordinated space)
str(crown_stem$matrix)
## And "series" the elements from this ordinated space into the two sub-spaces
str(crown_stem$series)
## Note that the elements here are indicated as numeric values referring to the row numbers in the
## original matrix
```

@@

```{r}
## Adding all mammals to the a new series (along with crown and stem as before)
series_factors2 <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem)),
    rep("Mammalia", Ntip(BeckLee_tree))), ncol = 2,
    dimnames = list(c(crown, stem), c("Group", "Class")), byrow = FALSE))

## Creating the three series: Group.stem/crown and Class.Mammalia
cust.series(BeckLee_mat50, series_factors2)
```

@@

Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
The `time.series` function allows to create series containing all the elements present at specific points in time or during specific periods.
This functions needs as input an ordinated space and a matching phylogenetic tree.
Two types of time series can be performed by using the `method` option:

@@

For the time-slicing method details see @GuillermeSTD.
<!-- TODO: link to the time-slicing algorithm explained! -->
For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:

@@

```{r}
## Generating three time bins containing the data present every 40 Ma
time.series(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = c(120, 80, 40, 0))
```

In this example, the taxa were split inside each time-bin according to their age.
Note however that some taxa might span between some time bins and should be included in more than one.
This is possible by providing a table containing the first and last occurrence data:

```{r}
## Generating three time bins containing the data present every 40 Ma, including taxa that might
## span between time bins.
(time_bins <- time.series(data = BeckLee_mat50, tree = BeckLee_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
```

This generated indeed a `dispRity` object of 3 series.
Note that we can also generate equivalent results by just telling the function that we want three time-bins (series) as follow:

@@

For the time-slicing method (`method = continuous`), the idea is fairly similar.
This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument):

@@

```{r}
## Generating four time slices every 40 million years assuming a gradual evolutionary model
(time_slices <- time.series(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages))
## Note that in the same way as before, we can also automatically generate the slices
time.series(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
    model = "gradual", time = 4)
```

@@

Once we obtain our different series, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each series more robust to outliers and the rarefaction will allow us to compare series with the same number of elements to get rid of eventual sampling problems.
The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
The default options will bootstrap the matrix 1000 times without rarefaction using the "full" bootstrap method (see below).

```{r}
boot.matrix(data = BeckLee_mat50)
```

The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option.
The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
Currently two algorithms are implemented:

@@

This function also allows to rarefy the data using the `rarefaction` argument.
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each series to a minimum of 3 elements).
It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
When left missing, all dimensions from the ordinated matrix are used.

@@

## Removing the last axis 50% last dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

## Keeping the 10 first dimensions
boot.matrix(BeckLee_mat50, dimensions = 10)
```

Of course, one could be interested in directly supplying the series generated above directly to this function.
In fact, it can also deal with a list of matrices or with a `dispRity` object output from the `cust.series` or `time.series` functions.

```{r}
## Bootstrapping and fully rarefying the crown/stem series
(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))

## Bootstrapping on the time binning/slicing series
(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
```

@@

This function is a modular function that allows to simply (and quickly!) calculate disparity from a matrix.
Because disparity can be measured in many ways, this function is a tool to measure disparity *as defined by the user*.
In fact, the `dispRity` function intakes two main arguments: the data and the disparity metric.
The disparity metric is a function or a set of functions that summarises the ordinated matrix to a single value that represents, in our example, the diversity of morphologies.

@@

```{r}
## Calculating disparity as the mean value of the ordinated space (using base::mean)
dispRity(BeckLee_mat50, metric = mean)

## Calculating disparity as the sum of variances of the ordinated space using base::sum on the
## variances and dispRity::variances on each dimension of the ordinated spaces
dispRity(BeckLee_mat50, metric = c(sum, variances))

## Calculating disparity as the product of the centroids (using base::prod and dispRity::centroids)
dispRity(BeckLee_mat50, metric = c(prod, centroids))

## Defining the total range function:
total.range <- function(X) abs(range(X)[1]-range(X)[2])

## Calculating disparity as the total range of the centroids:
dispRity(BeckLee_mat50, metric = c(total.range, centroids))
```

Note that in practice, the order of the functions in the `metric` argument does not matter:

```{r}
## Disparity as the standard deviation of the variances of the variance/covariance matrix;
vars1 <- dispRity(BeckLee_mat50, metric = c(sd, variances, var))

## Same disparity metric but with the functions ordered differently
vars2 <- dispRity(BeckLee_mat50, metric = c(variances, var, sd))

## Both versions gives the same results
all(summary(vars1) == summary(vars2))
```

@@

In these examples we used only simple ordinated matrix but of course, it possible to directly use the `dispRity` objects we generated in the steps above.
Thus we can calculate the bootstrapped and rarefied disparity in each series of the ordinated space.
Here we are going to define disparity as being the median distance between each element and each dimensions' centre.

@@

## Disparity through time:
disp_time_binsBS <- dispRity(time_binsBS, metric = c(median, centroids))
disp_time_slicesBS <- dispRity(time_slicesBS, metric = c(median, centroids))
```

Note that the computational time was longer for the rarefied data (`crown_stemBS`) since it had to calculate the disparity on all the rarefied matrix, each bootstrapped 100 times.
For both the `boot.matrix` and the `dispRity` functions, it is possible to increase computational speed in some cases using the `parallel` option (see each [parallel section](#parallel) below).

@@

This function is a S3 function (`summary.dispRity`) allowing to summarize the content of `dispRity` objects that contain disparity calculations.

```{r}
summary(disp_time_binsBS)
```

This function intakes a `dispRity` object plus various options namely the `quantile` values for the confidence intervals levels; the `cent.tend` for the 
central tendency to use for summarising the results and two *visual* options which are whether to recall the `dispRity` options and how much digits are wanted in the results.
By default the 50 and 95 quantiles and the median are calculated but these options can be modified as follows:

<!-- STOPPED HERE -->

```{r}
## Summarising the disparity using the 88 quantile and the standard deviation as the central
## tendency. 
summary(disp_time_binsBS, quantile = 88, cent.tend = sd)

@@

### `dispRity` object utilities  <a name="dispRity.utilities"></a>

export(make.dispRity)
export(fill.dispRity)
export(matrix.dispRity)
export(get.series.dispRity)
export(extract.dispRity)
export(scale.dispRity)
export(sort.dispRity)

### Package general utilities

## `tree.age`

@@

parallel  <a name="parallel"></a>
---------

Some of the core functions from the package have a `parallel` option.

Suggestions
============
If you have any extra suggestions or comments on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).
